Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-10-29 paper/paper.tex

\section{Introduction}
%% \fixme{Huang and those who reference him use the terms 'numerical
%%   deterministic reaction-diffusion model' and 'stochastic version of
%%   this model' for the stochastic.}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in preventing
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which itself interacts with another protein,
MinE, in such a fashion as to exhibit pole to pole oscillatory
behavoir.~\cite{hu1999topological, fu2001mine, shapiro2009and,
  yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

@@

There are only on the order of 1000 MinD proteins in a given cell,
which means that stochastic fluctuation can be expected to play a
large role.  To treat this, there have been a number of variations of
the Huang 2003 model that stochastically treat molecular interaction
and diffusion~\cite{fange2006noise, kerr2006division}\fixme{talk about
  this awkward sentence}.  These studies largely confirm the results
of Huang's model when applied to the wild-type, pill-shaped phenotype.
However, the stochastic models more successfully predict
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and enable prediction
of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

\begin{multline}
  \frac{\partial \rho_{D:ADP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ADP}-k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  +\delta(d_w)k_{de}\sigma_{DE},\hspace{3.4cm}
\end{multline}
\begin{multline}
  \frac{\partial \rho_{D:ATP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ATP}+k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  -\delta(d_w)[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
\end{multline}
\begin{multline}
  \frac{\partial \rho_E}{\partial t} = \mathcal{D}_E\nabla^2\rho_E+\delta(d_w)k_{de}\sigma_{DE}
  -\delta(d_w)k_E \sigma_D \rho_E
\end{multline}
\begin{multline}
  \frac{\partial \sigma_D}{\partial t} = -k_E\sigma_D\rho_E
  +[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
  \label{eq:d-on-wall}
\end{multline}
\begin{multline}
  \frac{\partial \sigma_{DE}}{\partial t} = -k_{de}\sigma_{DE}+k_E\sigma_D\rho_E\hspace{3cm}
  \label{eq:FifthPDE}
\end{multline}
where $\rho$ is cytoplasmic protein density ($\micron^{-3}$), $\sigma$
is membrane bound density ($\micron^{-2}$), $\mathcal{D}_D$ and
$\mathcal{D}_{E}$ are the diffusions constants for MinD and MinE,
respectively, $k_D^{\textrm{ADP $\rightarrow$ ATP}}$ is the rate of
conversion from MinD:ADP to the MinD:ATP complex, $k_D$ is the rate of
MinD:ATP attachement to the membrane when no protein is already
attached there, $k_{dD}$ is the increase of this rate when MinD:ATP is
present on the membrane, $k_{de}$ is the rate of hydrolisis of the
MinD:MinE:ATP complex, $k_E$ is the rate of cytoplasmic MinE binding
to membrane bound MinD:ATP complex, and $d_w$ is the distance from the
point in space to the closest wall.  The Dirac delta function
$\delta(d_w)$, which we need to describe the location of the membrane,
has units of $\micron^{-1}$ and is zero everywhere except at the wall.
Equations \ref{eq:d-on-wall} and \ref{eq:FifthPDE} are only relevent
at the membrane because the membrane-bound density values have no
meaning in the cytoplasm.

@@

Huang \emph{et al.}~\cite{huang2003dynamic} has performed a linear
stability analysis on the model which shows an upper limit on a steady
state solution of a $2\micron$ half wavelength.  Cells with dimensions
longer than this show spontaneous oscillatory behavoir in this
direction, while cells that are shorter relax into a motionless steady
state.  We have similarly performed a stability analysis in an
infinite slab with a thickness equal to that of our flattened cells
($0.25\micron$) and have found that an equivalent stability limit for
half wavelength is $2.13\micron$. As expected, when decreasing the
lengths and widths of our simulated flattened cells so that the
longest distance across the cell is less than this length, the cells
stop exhibiting any oscillatory behavior.  The deterministic model
relaxes into a motionless state and the stochastic model exhibits
random fluctuations without spatial oscillations.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the pill-shaped cell, shown against
    the correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  The stochastic model shows an
    oscillation period of 39.5 seconds and decay time of 307 seconds.}
  \label{corr-pill}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{A} (above) and the
    stadium \emph{A} (below) cell shapes, shown against the
    correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  For shape \emph{A}, the stochastic
    model shows an oscillation period of 53.1 seconds and decay time
    of 246 seconds, so that it takes roughly 4.6 periods for the
    behavoir to decohere.  For the stadium \emph{A}, the same model
    shows an oscillation period of 51 seconds and decay time of 323
    seconds, so that it takes roughly 6.3 periods for the behavoir to
    decohere.}
  \label{fig:corr-pancake-A}
\end{figure}

@@

We see that both in the shape \emph{A} and stadium \emph{A}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against eachother.  At
$\tau=0$, the stochastic correlation is already seen to be lower than
the deterministic, reflecting the fact that the stochastic data
sometimes shows maxima or minima at both ends simultaneously.  The
stochastic simulation of shape \emph{A} has a coherence time of 4.6
periods, while the corresponding simulation of stadium \emph{A} has a
coherence time of 6.3 periods.  This moderate difference in coherence
time shows that the regularity in cell oscillation is not overly
sensitive to the regularity and symmetry of the shape.  It seems that
it is the size and aspect ratio that are more important in dictating
the regularity of bipolar oscillation.

>>>>

\section{Introduction}
%% \fixme{Huang and those who reference him use the terms 'numerical
%%   deterministic reaction-diffusion model' and 'stochastic version of
%%   this model' for the stochastic.}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in preventing
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which itself interacts with another protein,
MinE, in a system which exhibits pole-to-pole oscillatory
behavoir~\cite{hu1999topological, fu2001mine, shapiro2009and,
  yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

@@

There are only on the order of 1000 MinD proteins in a given cell,
which means that stochastic fluctuation can be expected to play a
large role.  To treat this, there have been a number of variations of
the Huang 2003 model that stochastically simulate the same
reaction-diffusion equations~\cite{fange2006noise, kerr2006division}.
These studies largely confirm the results
of Huang's model when applied to the wild-type, pill-shaped phenotype.
However, the stochastic models more successfully predict
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and enable prediction
of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

\begin{multline}
  \frac{\partial \rho_{D:ADP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ADP}-k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  +\delta(d_w)k_{de}\sigma_{DE},\hspace{3.4cm}
\end{multline}
\begin{multline}
  \frac{\partial \rho_{D:ATP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ATP}+k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  -\delta(d_w)[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
\end{multline}
\begin{multline}
  \frac{\partial \rho_E}{\partial t} = \mathcal{D}_E\nabla^2\rho_E+\delta(d_w)k_{de}\sigma_{DE}
  -\delta(d_w)k_E \sigma_D \rho_E
\end{multline}
\begin{multline}
  \frac{\partial \sigma_D}{\partial t} = -k_E\sigma_D\rho_E
  +[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
  \label{eq:d-on-wall}
\end{multline}
\begin{multline}
  \frac{\partial \sigma_{DE}}{\partial t} = -k_{de}\sigma_{DE}+k_E\sigma_D\rho_E\hspace{3cm}
  \label{eq:FifthPDE}
\end{multline}
where $\rho$ is cytoplasmic protein density (proteins$/\micron^{3}$), $\sigma$
is membrane bound density (proteins$/\micron^{2}$), $\mathcal{D}_D$ and
$\mathcal{D}_{E}$ are the diffusions constants for MinD and MinE,
respectively, $k_D^{\textrm{ADP $\rightarrow$ ATP}}$ is the rate of
conversion from MinD:ADP to the MinD:ATP complex, $k_D$ is the rate of
MinD:ATP attachement to the membrane when no protein is already
attached there, $k_{dD}$ is the increase of this rate when MinD:ATP is
present on the membrane, $k_{de}$ is the rate of hydrolisis of the
MinD:MinE:ATP complex, $k_E$ is the rate of cytoplasmic MinE binding
to membrane bound MinD:ATP complex, and $d_w$ is the distance from the
point in space to the closest wall.  The Dirac delta function
$\delta(d_w)$, which we need to describe the location of the membrane,
has units of $\micron^{-1}$ and is zero everywhere except at the wall.
Equations \ref{eq:d-on-wall} and \ref{eq:FifthPDE} are only relevent
at the membrane because the membrane-bound density values have no
meaning in the cytoplasm.

@@

Huang \emph{et al.}~\cite{huang2003dynamic} has performed a linear
stability analysis on a cylindrical model which shows an upper limit
on a steady state solution of a $2\micron$ half wavelength.  Cells
with dimensions longer than this show spontaneous oscillatory behavoir
in this direction, while cells that are shorter relax into a
motionless steady state.  We have similarly performed a linear
stability analysis on an infinite slab with a thickness equal to that
of our flattened cells ($0.25\micron$) and have found that an
equivalent stability limit for half wavelength is $2.13\micron$. As
expected, when decreasing the lengths and widths of our simulated
flattened cells so that the longest distance across the cell is less
than this length, the cells stop exhibiting any oscillatory behavior.
The deterministic model relaxes into a motionless state and the
stochastic model exhibits random fluctuations without spatial
oscillations.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the pill-shaped cell, shown against
    the correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  The stochastic model shows an
    oscillation period of 39.5 seconds and decay time of 307 seconds.
    The correlation functions are scaled to have the same initial
    value.}
  \label{corr-pill}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{A} (above) and the
    stadium \emph{A} (below) cell shapes, shown against the
    correlation time.  Data for both the the determinisitic and
    stochastic models are shown, and the correlation functions are
    scaled to have the same initial value.  For shape \emph{A}, the
    stochastic model shows an oscillation period of 53.1 seconds and
    decay time of 246 seconds, so that it takes roughly 4.6 periods
    for the behavoir to decohere.  For the stadium \emph{A}, the same
    model shows an oscillation period of 51 seconds and decay time of
    323 seconds, so that it takes roughly 6.3 periods for the behavoir
    to decohere.}
  \label{fig:corr-pancake-A}
\end{figure}

@@

We see that both in the shape \emph{A} and stadium \emph{A}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against eachother.  The
stochastic simulation of shape \emph{A} has a coherence time of 4.6
periods, while the corresponding simulation of stadium \emph{A} has a
coherence time of 6.3 periods.  This moderate difference in coherence
time shows that the regularity in cell oscillation is not overly
sensitive to the regularity and symmetry of the shape.  It seems that
it is the size and aspect ratio that are more important in dictating
the regularity of bipolar oscillation.
